Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-18 paper/manuscript.tex
Wording tweaks

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}

@@

In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
\begin{equation}\label{eq:delta-delta-G-physical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
\end{split}
\end{equation}
Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
\begin{equation}\label{eq:delta-delta-G-alchemical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
\end{split}
\end{equation}
where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
Note that, when $A$ and $B$ have a different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.

@@

@@

\subsubsection{Changes in net charge can be challenging/problematic.}
\tocsubsubsectioncomment{@Chris Oostenbrink/ Vytas Gapsys to update.}
If the net charge of the system changes as the alchemical variable changes during the calculation, this can pose major challenges.
Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics, including particle mesh Ewald (PME) and reaction field, do not handle free energy contributions from such changes as they would be handled in a hypothetical macroscopic bulk solution. The outcome depends either on the size of the cutoff for reaction field methods or on the size of the simulation box when using lattice sum methods due to periodic boundary conditions~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.

Lattice summation methods (e.g. PME) effectively neutralize the simulation box by applying an implicit homogeneously distributed background counter charge~\cite{figueirido1995finite,hummer1996free}. The background charge can be interpreted as a converged distribution of counterions and may be a reasonable choice for computing hydration free energies of monovalent ions (finite size effect corrections can be applied in these cases)~\cite{hummer1996free}. In inhomogeneous systems with different dielectrics, e.g. solvated protein or lipid bilayer in a water box, assuming uniform background countercharge distribution is unphysical, as the higher countercharge density should be located in the regions of higher dielectric~\cite{hub2014pme}. This, in turn, causes significant artifacts in sampled population densities of charged molecules, i.e. errors in calculated free energy differences.

There are two main classes of approaches to address artefacts introduced by changes in net charge during alchemical free energy calculations: (1) avoiding changing the net charge altogether and (2) post-processing approaches to correct for artefacts using continuum electrostatics.
Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}. In absolute free energy calculations, it is often difficult to avoid a change in the net charge. One way to avoid charge changes is to ensure that, as a charged ligand is removed, a charged counterion of opposite sign is also removed, or one of the same sign is inserted. This is sometimes referred to as a “co-alchemical ion” approach to deal with the required charge change. This approach is easily supported in most free energy calculating programs, as it merely involves an additional simultaneous modification.
There are special flavours of the co-alchemical ion approach such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup. Here, the two branches of a thermodynamic cycle as used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation. Note that the co-alchemical approaches only avoid finite-size effects with lattice-sum methods such as PME.

Charge-change correction schemes have been explored and offer potentially viable solutions for the problem~\cite{mey2018impact}, where artifacts introduced by finite-size effects are corrected a posteriori from Poisson-Boltzmann calculations ~\cite{chen2018accurate, ohlknecht2020correcting, rocklin2013calculating,Reif2014chargecorrection}. However, application of such corrections is typically inconvenient and computationally demanding. High grid resolutions for the Poisson-Boltzmann calculations are required to avoid numerical artifacts, and it may be necessary to average over multiple representative configurations to obtain reliable results. Furthermore, they rely on continuum electrostatic models, which cannot fully capture phenomena arising from discrete solvent effects. These effects can be quantified in the orientational-disorder limit such that analytical correction terms for this effect exist for solvent models with a single van der Waals interaction site. Readers interested in these conceptual and technical limitations are referred to detailed analyses by Hünenberger and co-workers~\cite{kastenholz2006solvfe, rocklin2013calculating}.
Alternatively, Reif and Oostenbrink demonstrated in a proof-of-concept study that on-the-fly corrections to the physical forces can be effectively used to eliminate finite-size and approximate electrostatic artifacts~\cite{Reif2015onthefly}.

Petrov et al.~\cite{Petrov2024guidelines} investigated the box-size dependencies of alchemical free energy calculations using PME together with the co-alchemical ion approach. They concluded that in neutral simulation boxes with a minimal distance to the box wall larger than 1 nm, finite-size artifacts could be made negligible. The following recommendations should be followed to minimize finite-size artifacts, when using lattice-sum electrostatics: (a) setup of free energy calculations such that there is no net charge change whenever possible, (b) ensuring an overall neutral simulation box, (c) use of a sufficiently large distance between the solute and the box wall, at least 1 nm, and (d) use of salt if possible to screen finite-size periodicity artifacts~\cite{chen2018accurate}. In cases where these recommendations are not practical for specific applications, such as small simulation boxes or when reaction field methods are preferred, charge-change corrections may be applied to reduce artifacts. Though both post-processing corrections and the co-alchemical ion approach come at a cost of increased uncertainties. The Poisson-Boltzmann calculations are approximate due to finite-grid sizes and are applied only to a subset of the ensemble. Co-alchemical ion approaches introduce additional degrees of freedom and increase sampling requirements.

@@

\subsubsection{The importance of configurational sampling}
\label{sec:configurational_sampling}
As mentioned previously, free energy differences estimated from alchemical calculations should be independent of the initial system conformation, given sufficient configurational sampling. In cases where the system's slowest degrees of freedom are coupled with the alchemical variable, transitioning between alchemical states can indirectly enhance configurational sampling. This effect is particularly pronounced when certain alchemical states lack the energy barriers present at the fully coupled state (e.g. $\lambda=0$ in Fig. ~\ref{fig:configurational_sampling}\textbf{A}), thereby allowing the system to bypass otherwise prohibitive configurational free energy barriers. For example, in a protein-ligand unbinding process where the kinetic barrier is predominantly associated with the disruption of intermolecular interactions or the separation of binding partners, simulating different intermediate states effectively samples configurations with varying center-of-mass distances, thus mitigating the sampling issue in the configurational space.

Conversely, when the slowest degrees of freedom in the configurational space are largely orthogonal to the alchemical variable, the system can be stuck in its initial metastable state. Such entrapment leads to discrepancies in free energy estimates from simulations initiated from different conformations. This scenario, as illustrated in Fig.~\ref{fig:configurational_sampling}\textbf{B}, is not uncommon especially in flexible systems. For example, it can arise when the free energy difference of interest needs to account for metastable states separated by torsional barriers, such as the solvation free energy of a ligand having different rotamer states, the methylation free energy of a nucleobase in a duplex ~\cite{hsu2023alchemical}, and the binding free energy that involves flexible side chains near the binding site~\cite{la2022alchemical}. Similar sampling challenges have also been reflected in several recent studies, which highlights the sensitivity of alchemical calculations to initial conformations, whether derived from crystallographic structures~\cite{suruzhon2021sensitivity, baumann2024impact} or docking protocols~\cite{cappel2020impact}.

Similarly, Fig.~\ref{fig:configurational_sampling}\textbf{C} highlights the importance of configurational sampling in alchemical calculations, where a free energy basin spans both the alchemical and configurational coordinates, potentially trapping the system in both dimensions. This can occur, for instance, when electrostatic interactions are turned off while van der Waals interactions remain, rendering the alchemically transformed molecule a ``greasy ball''. Notably, in cases where alchemical biases/weights are applied to facilitate even alchemical sampling across states, such as in an expanded ensemble simulation~\cite{lyubartsev1992new}, scenario Fig.~\ref{fig:configurational_sampling}\textbf{C} may evolve toward scenario Fig.~\ref{fig:configurational_sampling}\textbf{B} as the alchemical free energy profile is flattened out, while the basin in the configurational direction persists.

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach involves performing separate alchemical calculations from different metastable states, then computing the overall free energy difference as the negative log-sum-exp of the Boltzmann weights from each metastable state~\cite{hsu2023alchemical}. However, this method is only applicable to cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}, where the system remains trapped in its initial metastable state throughout the calculation. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from different metastable states) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda} and its combination~\cite{zhou2025zooming} with well-tempered metadynamics~\cite{barducci2008well}, to name a few.

Notably, detecting and addressing configurational sampling challenges in alchemical calculations remains a non-trivial task. In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}. In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them. Such collective variables are often based on physical intuition, experimental insights, or preliminary computational analyses. However, optimal collective variables are rarely known \textit{a priori}, complicating the systematic detection of sampling issues. Although recent ad-hoc machine learning methods for optimal CV identification have shown promise~\cite{gokdemir2025machine, sidky2020machine, bonati2021deep} and could in principle be straightforwardly integrated into alchemical sampling protocols, systematic studies explicitly demonstrating such integration have yet to appear. Consequently, widely accepted best practices about automated diagnostic protocols capable of identifying and resolving configurational sampling bottlenecks have not yet been established.

\begin{figure}
    % Figure to be updated
    \includegraphics[width=0.88\columnwidth]{paper/figures/fig_configurational_sampling/FES_scenarios_path.pdf}
    \caption{\textbf{Different scenarios of free energy surfaces.} In scenario \textbf{A}, the free energy barrier present at $\lambda=0$ is absent at other $\lambda$ states, so the system can go around the barrier by alternating sampling between alchemical and configurational space, e.g. via the path shown as the dashed line. In contrast, in scenario \textbf{B}, the free energy barrier extending all alchemical states prevents the system from sampling both metastable states at $\lambda=0$. In scenario \textbf{C}, the free energy basin extending in both alchemical and configurational directions can trap the system.}
    \label{fig:configurational_sampling}
\end{figure}

@@

The non-equilibrium switching (Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}) approach makes use of short out-of-equilibrium simulations~\cite{aldeghi2018accurate}. In this protocol, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states. 
Over the course of each transition, Hamiltonian derivative with respect to $\lambda$ is recorded; subsequently the curves are integrated to yield work distributions for the forward and reverse processes. Crooks Fluctuation Theorem~\cite{crooks1999entropy} allows estimating free energy difference between the end states from the generated work distributions.~\cite{shirts2003equilibrium}

The non-equilibrium (NEQ) approaches offer some advantages over the equilibrium sampling based methods. For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining ligand network: the transitions between the ligands can be generated afterwards. This eliminates the redundant equilibrium sampling, e.g. in a starmap network, FEP based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards. The NEQ protocol also provides flexibility in combining different structures in free energy calculations, e.g. simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates. Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.

The non-equilibrium approach is available in GROMACS and Amber. A closely related method, termed non-equilibrium cycling, is available in OpenMM. Among the commercial packages, it is implemented in OpenEye's Orion{\textregistered}~\cite{sorensen2024orion} platform.
%A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}.

@@

Consider, as an illustration, the binding equilibrium between a receptor $R$ and a ligand $A$ that exists in solution as a mixture of protonated $AH$ and ionized $A^-$ forms. It is often useful to consider the observed equilibrium binding constant between $R$ and the $A$ mixture 
\begin{equation}
  K_b(A) = C^\circ \frac{[RA]}{[R][A]}
  \label{eqn:observed-binding-constant}
\end{equation}
where $[A] = [AH] + [A^-]$ and $[RA] = [RAH] + [RA^-]$ are the total equilibrium concentrations of free $A$ and receptor-ligand complexes. The observed equilibrium constant is significant because it corresponds to binding affinity measurements that do not distinguish between the protonation states of the ligand. Inserting the definitions of $[A]$ and $[RA]$ in Eq.~(\ref{eqn:observed-binding-constant}), splitting the sum in the numerator, and multiplying and dividing each term by either $[AH]$ or $[A^-]$ we obtain
\begin{equation}
  K_b(A) = \frac{[AH]}{[A]}  K_b(AH) +  \frac{[A^-]}{[A]} K_b(A^-) =  p_0(AH) K_b(AH) +  p_0(A-) K_b(A^-)
  \label{eqn:observed-binding-constant-split}
\end{equation}
where
\begin{equation}
  K_b(AH) = C^\circ \frac{[RAH]}{[R][AH]}
  \label{eqn:observed-binding-constant-prot}
\end{equation}
is the specific equilibrium binding constant between the receptor and the protonated form of the ligand, and similarly for $K_b(A^-)$. The specific $K_b$'s are the binding constants that would be obtained from ABFE calculations for those species. Eq.~(\ref{eqn:observed-binding-constant-prot}) states that the observed binding constant is an average of the specific binding constants weighted by the populations, $p_0(AH)$ and  $p_0(A^-)$ of each ligand's chemical form in solution. In this specific example, if the $pKa$ of $AH$ is known and $\alpha = [A^-]/[AH] = 10^{pH-pKa} $ is the ionization ratio of $A$ at a a given $pH$, the populations of the ionized and protonated forms are $p_0(A^-) = \alpha/(\alpha+1)$ and $p_0(AH) = 1/(\alpha+1)$, respectively.\cite{champion2024multistate,azimi2022application}

@@

Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, non-reversible binding of organic molecules to well-structured proteins. However, several areas of drug discovery involve other classes of binding events. While the underlying theory of alchemical free energy calculations is generally applicable, additional work may be required to develop efficient protocols suited to systems beyond the traditional small molecule–protein complex.
For example, there is growing interest in identifying small molecules that directly bind to RNA or DNA, with recent preliminary studies showing promising results using specialized simulation protocols.\cite{Abramyan2025,Rasouli2024} In contrast, there has been limited investigation into the applicability of AFE methods to small-molecule ligands targeting intrinsically disordered proteins (IDPs,\cite{Papadourakis2024} or proteins containing intrinsically disordered regions (IDRs) that undergo disorder-to-order transitions upon ligand binding.\cite{MendozaMartinez2022}
The treatment of covalent binders has also received some attention.\cite{Luo2021,Zhang2019 } However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019}

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[11]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}

@@

In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
\begin{equation}\label{eq:delta-delta-G-physical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
\end{split}
\end{equation}
Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
\begin{equation}\label{eq:delta-delta-G-alchemical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
\end{split}
\end{equation}
where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
Note that, when $A$ and $B$ have a different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.

@@

\paragraph{Membrane proteins require additional care}
Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations. Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}). Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}. These systems require more extensive equilibration than soluble proteins. Some workflows, such as CHARMM-GUI or Desmond membrane preparation, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.

Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water. Several approaches are reviewed in \cite{wu2022correction-51c}. The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.

In the case of a lipid-facing binding site, there are three additional challenges. First, simulation time should be extended to ensure sufficient sampling of lipid configurational space. Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand. If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols. Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids. Second, considering the free energy cost of a ligand partitioning into the membrane may provide additional insights \cite{doi:10.1021/acs.jcim.1c01147}. Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species. In some cases, the identity of the lipid may be known. But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.

@@

\subsubsection{Changes in net charge can be challenging/problematic.}
\tocsubsubsectioncomment{@Chris Oostenbrink/ Vytas Gapsys to update.}
If the net charge of the system changes as the alchemical variable changes during the  calculation, this can pose major challenges.
Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.

There are two main potential solutions to avoid artifacts due to changes in net charge: avoiding changing the net charge, and correcting for the introduced artifacts.

Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}. Absolute free energy calculations can also potentially avoid changing the charge of the system by making a charge perturbation of equal and opposite sign elsewhere in the system; for example, as a charged ligand is removed, a charged counterion of opposite sign could also be removed, or one of the same sign could be inserted. This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate, ohlknecht2020correcting}. However, application of such corrections typically remains less common than the use of a co-alchemical ion. A third approach has been proposed by Gapsys et al.~\cite{gapsys2015calculation} which uses a double-system/single box setup.

When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem, and several papers give good guides for exploring the problem further~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.

@@

@@

The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}. In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states. This approach is available in GROMACS and appears to be coming online in several other packages. A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}.

@@

Consider, as an illustration, the binding equilibrium between a receptor $R$ and a ligand $A$ that exists in solution as a mixture of protonated $AH$ and ionized $A^-$ forms. It is often useful to consider the observed equilibrium binding constant between $R$ and the $A$ mixture 
\begin{equation}
  K_b(A) = C^\circ \frac{[RA]}{[R][A]}
  \label{eqn:observed-binding-constant}
\end{equation}
where $[A] = [AH] + [A^-]$ and $[RA] = [RAH] + [RA^-]$ are the total equilibrium concentrations of free $A$ and receptor-ligand complexes. The observed equilibrium constant is significant because it corresponds to binding affinity measurements that do not distinguish between the protonation states of the ligand. Inserting the definitions of $[A]$ and $[RA]$ in Eq.~(\ref{eqn:observed-binding-constant}), splitting the sum in the numerator, and multiplying and dividing each term by either $[AH]$ or $[A^-]$ we obtain
\begin{equation}
  K_b(A) = \frac{[AH]}{[A]}  K_b(AH) +  \frac{[A^-]}{[A]} K_b(A^-) =  p_0(AH) K_b(AH) +  p_0(A-) K_b(A^-)
  \label{eqn:observed-binding-constant-split}
\end{equation}
where
\begin{equation}
  K_b(AH) = C^\circ \frac{[RAH]}{[R][AH]}
  \label{eqn:observed-binding-constant-prot}
\end{equation}
is the specific equilibrium binding constant between the receptor and the protonated form of the ligand, and similarly for $K_b(A^-)$. The specific $K_b$'s are the binding constants that would be obtained from ABFE calculations for those species. Eq.~(\ref{eqn:observed-binding-constant-prot}) states that the observed binding constant is an average of the specific binding constants weighted by the populations, $p_0(AH)$ and  $p_0(A^-)$ of each ligand's chemical form in solution. In this specific example, if the $pKa$ of $AH$ is known and $\alpha = [A^-]/[AH] = 10^{pH-pKa} $ is ionization ratio of $A$ at a a given $pH$, the populations of the ionized and protonated forms are $p_0(A^-) = \alpha/(\alpha+1)$ and $p_0(AH) = 1/(\alpha+1)$, respectively.\cite{champion2024multistate,azimi2022application}

@@
